---
title: "Air quality model benchmarking with _dartle_"
author: "Giovanni Bonafè"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Air quality model benchmarking with dartle}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


The R package _dartle_ is a toolkit of functions for air quality model benchmarking, inspired by the [DELTA tool](http://aqm.jrc.ec.europa.eu/index.aspx) (JRC-IES) and the work of the [FAIRMODE](http://fairmode.jrc.ec.europa.eu/) WP1.

For _dartle_ to work, you need `R` (version 2.10 or higher) and the packages `stats`, `dplyr`, `RcppRoll`, `data.table` and `ggplot2`. To install _dartle_ downloading it from GitHub, use the package `devtools`, as follows:

```r
require("devtools")
devtools::install_github("jobonaf/dartle")
```

Once installed, you can load it
```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
library("dartle")
```


## Datasets

The package includes some datasets:


|Item                |Title                         |
|:-------------------|:-----------------------------|
|mod.data (mod_data) |Forecasted concentrations     |
|obs.no2 (obs_data)  |Observed NO2 concentrations   |
|obs.o3 (obs_data)   |Observed ozone concentrations |
|obs.pm10 (obs_data) |Observed PM10 concentrations  |
|obs.pm25 (obs_data) |Observed PM2.5 concentrations |

Let's have a look to the forecasted data (ignore `EmissionTime`, `Intercept` and `Slope`)
```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
library(knitr)
kable(head(mod.data))
```

Let's see where and when have been observed the highest concentrations. Of PM10
```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
require(dplyr)
kable(head(obs.pm10 %>% arrange(desc(Value))))
```


PM2.5
```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
kable(head(obs.pm25 %>% arrange(desc(Value))))
```


NO2
```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
kable(head(obs.no2 %>% arrange(desc(Value))))
```


and ozone
```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
kable(head(obs.o3 %>% arrange(desc(Value))))
```

Note that for PM10, NO2 and ozone only background stations are provided, while `obs.pm25` includes also traffic and industrial stations. All the data are from the [Environmental Agency](http://www.arpa.fvg.it) of Friuli Venezia Giulia region (Italy).
```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
unique(obs.pm25$StationType)
```

## Target plots

To produce a _target plot_, we must arrange forecasted and observed data in the same _data.frame_. First, we extract PM10 forecasts with `filter` and calculate the daily averages with `dMean`
```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
Mod <- dMean(mod.data %>% filter(Var=="c_PM10"),
             value = "Value", time = "Time", point = "Point")
```

Then, we adjust the format of the observed dataset
```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
Obs <- obs.pm10 %>% mutate(Day=format(Time,"%Y-%m-%d"), Point=ID)
```

So, observed and forecasted datasets can be meld in a single _data.frame_ (`Dat.pm10`)
```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
Dat.pm10 <- inner_join(Mod, Obs, by=c("Point", "Day"), suffix = c(".mod", ".obs"))
```

Now `Dat` can be passed to the `target_report` function, to calculate some quality indicators
```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
t_rep <- target_report(Dat.pm10, obs = "Value.obs", mod = "Value.mod",
                       point = "Point", pollutant = "PM10")
```

Finally, `t_rep` (the output of the `target_report` function) is ready to be passed to the `target_plot` function
```{r, fig.show='asis', fig.height=7, fig.width=6, message=FALSE, warning=FALSE} 
target_plot(t_rep)
```

The same processing for PM2.5:
```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
Mod <- dMean(mod.data %>% filter(Var=="c_PM25"),
             value = "Value", time = "Time", point = "Point")
Obs <- obs.pm25 %>% mutate(Day=format(Time,"%Y-%m-%d"), Point=ID)
Dat.pm25 <- inner_join(Mod, Obs, by=c("Point", "Day"), suffix = c(".mod", ".obs"))
t_rep <- target_report(Dat.pm25, obs = "Value.obs", mod = "Value.mod",
                       point = "Point", pollutant = "PM2.5")
```

And the plot
```{r, fig.show='asis', fig.height=7, fig.width=6, message=FALSE, warning=FALSE} 
target_plot(t_rep)
```

For NO2, the target plot is based on hourly data, therefore no daily average is performed
```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
Mod <- mod.data %>% filter(Var=="c_NO2") %>% mutate()
Obs <- obs.no2 %>% rename(Point=ID)
Dat.no2 <- inner_join(Mod, Obs, by=c("Point", "Time"), suffix = c(".mod", ".obs"))
t_rep <- target_report(Dat.no2, obs = "Value.obs", mod = "Value.mod",
                       point = "Point", pollutant = "NO2")
```

Note that the function `target_report` needs the argument `pollutant` because the parameters used to assess the measurements uncertainties are pollutant-dependent. 
```{r, fig.show='asis', fig.height=7, fig.width=6, message=FALSE, warning=FALSE} 
target_plot(t_rep)
```


For ozone, we have to calculate daily maxima of the 8-hours running mean. 
```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
Mod <- dMaxAvg8h(mod.data %>% filter(Var=="c_O3"),
             value = "Value", time = "Time", point = "Point")
Obs <- obs.o3 %>% rename(Point=ID)
```
In this case, both observations and forecasts are provided as hourly averages in the datasets, so we need the function `dMaxAvg8h` again.
```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
Obs <- dMaxAvg8h(Obs, value = "Value", time = "Time", point = "Point")
Dat.o3 <- inner_join(Mod, Obs, by=c("Point", "Day"), suffix = c(".mod", ".obs"))
t_rep <- target_report(Dat.o3, obs = "Value.obs", mod = "Value.mod",
                       point = "Point", pollutant = "O3")
```


```{r, fig.show='asis', fig.height=7, fig.width=6, message=FALSE, warning=FALSE} 
target_plot(t_rep)
```

## Summary reports

```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE, fig.height=7, fig.width=6}
s_rep <- summary_report(Dat.pm10, obs = "Value.obs", mod = "Value.mod",
                       point = "Point", pollutant = "PM10")
summary_plot(s_rep)
```


```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE, fig.height=7, fig.width=6}
s_rep <- summary_report(Dat.pm25, obs = "Value.obs", mod = "Value.mod",
                       point = "Point", pollutant = "PM2.5")
summary_plot(s_rep)
```


```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE, fig.height=7, fig.width=6}
s_rep <- summary_report(Dat.no2, obs = "Value.obs", mod = "Value.mod",
                       point = "Point", pollutant = "NO2")
summary_plot(s_rep)
```



```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE, fig.height=7, fig.width=6}
s_rep <- summary_report(Dat.o3, obs = "Value.obs", mod = "Value.mod",
                       point = "Point", pollutant = "O3")
summary_plot(s_rep)
```


## References
* [Guidance Document on Modelling Quality Objectives and Benchmarking](http://fairmode.jrc.ec.europa.eu/document/fairmode/WG1/Guidance_MQO_Bench_vs2.1.pdf)  

----
<sub>
<div>Logo made by <a href="http://www.freepik.com" title="Freepik">Freepik</a> from <a href="http://www.flaticon.com" title="Flaticon">www.flaticon.com</a> is licensed by <a href="http://creativecommons.org/licenses/by/3.0/" title="Creative Commons BY 3.0" target="_blank">CC 3.0 BY</a></div>
</sub>
